---
title: Lab Ten -- Point Estimation # Title of the document

# Bibliography Details
bibliography: Lab10References.bib  # Name of bib file in the same folder 
bibliographystyle: apa            # Bibliography style

execute:               # Default settings for code blocks
  warning: false       # Do not print warnings
  message: false       # Do not print messages
  size: "scriptsize"   # Set code size to script size

# Below will do html by default. To render the entire project
# Use the following code:
# quarto::quarto_render("Lab10WriteUp.qmd")
format:
  pdf:        # Specify .pdf formatting options 
    documentclass: article     # Document Class
    number-sections: true      # Number Sections
    reference-section-title: "References" # Title for reference section
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables 
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in
  html:       # Specify HTML formatting options 
    number-sections: true      # Number Sections
    code-fold: true            # Button to show code
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables and figures
format-links: [pdf]
---
```{r}
#| echo: false

default_chunk_hook  <- knitr::knit_hooks$get("chunk")

latex_font_size <- c("Huge", "huge", "LARGE", "Large", 
                     "large", "normalsize", "small", 
                     "footnotesize", "scriptsize", "tiny")

knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  if(options$size %in% latex_font_size) {
    paste0("\n \\", options$size, "\n\n", 
      x, 
      "\n\n \\normalsize"
    )
  } else {
    x
  }
})
```

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
library(patchwork)
```

**Tips**

* Complete the tasks below. Make sure to start your solutions in on a new line that starts with "**Solution**:".
* Make sure to use the Quarto Cheatsheet. This will make completing and writing up the lab *much* easier.
* Make sure to use "enter" to make your code readable, so it doesn't run off the page. I switched the Quarto document to automatically render to .pdf, so you can see what you're submitting by default. You can switch back for the highlighting by moving the html block above the pdf block in the YAML header.
* Don't forget that you can copy-and-paste code when you're doing something similar as we did in class!
* Make sure to plot all computed probabilities. This is good `ggplot` practice AND will help make sure you don't make a mistake when computing the probability.


# Lab 8 Notes

##  Altham's Multiplicative Binomial Distribution

The PMF for Altham's Multiplicative Binomial Distribution is
$$f_X(x) = \frac{1}{c} {n \choose x} p^x (1-p)^{n-x} \theta^{x(n-x)}I(x \in \{0, 1, ..., n\})$$
where
$$c = \sum_{i=0}^n {n \choose i} p^i (1-p)^{n-i} \theta^{i(n-i)}.$$ 
Just so we start on the same page, I have included my `R` function for computing probability using this distribution.

```{r}
daltbinom <- function(x, size, prob, theta){
  normalizing.constant <- 0
  for(j in 0:size){
    normalizing.constant <- normalizing.constant + 
      choose(size, j) * prob^j * (1-prob)^(size-j) * theta^(j*(size-j))
  }
  # Probability Mass at x=x
  (1/normalizing.constant) *
    choose(size, x) * prob^x * (1-prob)^(size-x) * theta^(x*(size-x)) *
      (x>=0)*(x<=size)
}
```

# Altham's Multiplicative Beta Binomial Distribution
The PMF for Altham's Multiplicative Beta Binomial Distribution is
$$f_X(x) = \left[\int_{0}^1 \left(\frac{1}{c} {n \choose x} p^x (1-p)^{n-x} \theta^{x(n-x)}\right)\left(\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)} p^{\alpha-1}(1-p)^{\beta-1}\right) dp\right]I(x \in \{0, 1, ..., n\})$$

Just so we start on the same page, I have included my `R` function for computing probability using this distribution.
```{r}
daltbbinom.single <- function(x, size, alpha, beta, theta){
    integrand <- function(p) {
    sapply(p, # integrate will pass in a vector, so we use sapply to apply the following
           function(p_val){
             # f(x|p) * f(p|alpha, beta)
             daltbinom(x, size, p_val, theta) * dbeta(p_val, alpha, beta)
           }
    )
  }
  # Probability Mass at x=x
  # integrate(integrand, lower = 0, upper = 1)$value * (x>=0)*(x<=size)
  # I shrink the bounds below to avoid log(0)=-infinty
  # I use try() to catch when there is an error and avoid crashing
  res <- try(integrate(integrand, lower = 1e-7, upper = 1 - 1e-7)$value * 
                  (x>=0)*(x<=size), 
             silent = TRUE)
  # If there is an error, return a near-zero probability
  if(inherits(res, "try-error")) return(0.1^6)
  
  res
}
# Write a function that works on a vector
daltbbinom <- function(x, size, alpha, beta, theta){
  sapply(X=x, FUN=daltbbinom.single, size=size, alpha=alpha, beta=beta, theta=theta)
}
```

# Question 1

In Lab 8, we use Altham's Multiplicative Binomial Distribution. Your job was to interpret the impact of $\theta$ in that equation. In this lab, we will model the number of legs won in a best of eleven (first to six) match. 

To help us think about the interpretation, something that was challenging before, let's try something a little more manageable.

## Part a
Suppose a player has a 50\% chance of winning a specific leg, that is let $p=0.50$. Let's look at the probability they win ($x=6$) legs in the match. Compute this probability over a sequence from $\theta=0.5$ to $\theta=1.5$ and plot it using a line where the $x$-axis is $\theta$ and the $y$-axis is $P(X=6)$.

**Solution:**
```{r}
p = 0.5
n = 11
x = 6
theta = seq(0.5, 1.5, length.out = 300)

prob = sapply(theta, function(th){
  daltbinom(x, n, p, th)
})

df = tibble(theta = theta,
            prob = prob)

Q1PA = ggplot(df, aes(theta, prob)) +
              geom_line(color = "blue") +
              labs(title = "",
              x = expression(theta),
              y = "P(X = 6)") +
theme_bw()
Q1PA
```

## Part b
Now, plot Altham's Multiplicative Binomial Distribution under the same parameters with $\theta=0.50$, $\theta=1$, and $\theta=1.5$. Use `ncol=1` in `facet_wrap()` to plot the PMFs in a single column for comparing.
**Solution:**
```{r}
theta = c(0.5, 1, 1.5)
df = crossing(x = 0:n,
              Theta = theta) |>
              rowwise() |>
              mutate(prob = daltbinom(x, n, p, Theta)) |>
              ungroup() |>
              mutate(Theta = factor(Theta))

Q1PB = ggplot(df, aes(x = x, y = prob)) +
              geom_col(fill = "steelblue") +
              facet_wrap(~Theta, ncol = 1) +
              labs(title = "Altham's Multiplicative Binomial PMF",
              x = "Number of Legs Won (x)",
              y = "Probability") +
              theme_bw()
Q1PB
```

## Part c
What is your best assessment of what $\theta$ is doing in this setting. Consider when a 6-0 blowout is most likely, and when a 5-6 nail-biter is most likely. Note here we should treat the match as a fixed $n=11$ as we did for $n=3$ in Lab 8.\
**Note:** In lab 8, we saw that small $\theta$ meant wins came in bunches (clumped together), so there were more 2-0 wins in a best of three series.


**Solution:**
Perhaps small theta values make extreme scenarios more likely to happen, like 6-0 or 0-6 losses. On the other hand higher theta values make close games more likely to happen where you will see 5-6 or 6-5 record games.

# Question 2

## Part a
Altham's Multiplicative Binomial Distribution does not typically simplify to the clean $E(X)=np$ seen in the Binomial distribution, unless $\theta = 1$. Write out the expression that would need to be solved to find $E(X^k)$, and then write a function that computes it given the proposed parameter values $p$ (`prob`) and $\theta$ (`theta`).
**Solution:**
```{r}
exk = function(k, size, prob, theta){
              c = sum(sapply(0:size, function(x){
              choose(size, x) * prob^x * (1-prob)^(size-x) * theta^(x*(size-x))}))
  

              numerator = sum(sapply(0:size, function(x){
              x^k * choose(size, x) * prob^x * (1-prob)^(size-x) * theta^(x*(size-x))}))
  
              numerator / c
}
```

## Part b
Describe how you can use your answer to part a to find Method of Moments estimates for the parameters $p$ and $\theta$ for Altham's Multiplicative Binomial Distribution based on data.
**Solution:**
We can compute the sample mean and variance to find the average and spread of the data. We can then set these equal to the values attained from the "exk" funciton to solve for estimates of $p$ and $theta$.

## Part c
Write a function that computes the log-likelihood for Altham's Multiplicative Binomial Distribution, given data (`x`) and proposed parameter values $p$ (`prob`) and $\theta$ (`theta`).
**Solution:**
```{r}
loglike_altbinom = function(x, n, prob, theta){
  c = sum(sapply(0:n, function(i){
          choose(n, i) * prob^i * (1 - prob)^(n - i) * theta^(i * (n - i))}))
  
  loglike = sum(sapply(x, function(xi){
      log(choose(n, xi)) + xi * log(prob) + (n - xi) * log(1 - prob) +
      xi * (n - xi) * log(theta) - log(c)}))
  
  return(loglike)
}
```

## Part d
Describe how you can use your answer to part c to find Maximum Likelihood estimates for the parameters $p$ and $\theta$ for Altham's Multiplicative Binomial Distribution based on data.
**Solution:**
We can use the log-likelihood funciton to find maximums to $p$ and $theta$ by plugging in different values of $p$ and $theta$ until it returns a high enough value, or the peak of the funciton. This is the value combination that shows the highest likelihood of occur with the set $p$ and $theta$. 

# Lab 10
Peter "Snakebite" Wright is a two-time Professional Darts Corporation World Champion (2020, 2022) and a former World No. 1. He is one of the sport's most decorated players, winning nearly 50 PDC titles including the World Matchplay, UK Open, and multiple European Championships. 

In 2024, he made The Premier League -- a league involving the eight best players that runs every Thursday night from February to May. In this season, he won two of eighteen matches, securing only four points. The spread in legs won was -43, meaning he lost 43 more legs than he won. This performance was rather quite bad. He was soundly trounced in almost all his match ups. 


The 2024 standings are below:

| Rank | Player             |
|:-----|:-------------------|
| 1    |Luke Littler        |
| 2    |Luke Humphries      |
| 3    |Michael van Gerwen  |
| 4    |Michael Smith       |
| 5    |Nathan Aspinall     |
| 6    |Rob Cross           |
| 7    |Gerwyn Price        |
| 8    |Peter Wright        |


## Summarize the 2024 Data

In `pdc2024.csv`, I have provided the data for the 2024 Premier League season. There are a few data cleaning steps to consider.

  * Remove any matches with no `Score` (e.g., a bye or walkover)
  * Nights 1 through 16 are best of 11 (first to 6), but the playoffs are extended. For consistency, keep nights 1 through 16 only.
  * Make two columns `WinnerLegs` and `LoserLegs` that keep track of the number of legs each player has won
  
  Summarize the data. What do the data tell us about the breakdown of the results?
**Solution:**
The data shows that when players lose they don't get blown out, with the average loser still winning 3.56 legs and varying by 1.43. We also see that the #1 ranked player won a considerable about of games whereas the bottom ranked player won 0 games. This means the skill gap between the players is very high throughout the league.
```{r}
data = read.csv("pdc2024.csv") |>
        filter(Night != "Finals") |>
        filter(Score != "w/o") |>
        mutate(WinnerLegs = 6,
               LoserLegs = as.integer(substr(Score, nchar(Score), nchar(Score))))
top = 0
bottom = 0
for (player in data$Winner){
              if (player == "Luke Littler"){
                top = top + 1
              }
}
for (player in data$Winner){
              if (player == "Peter Wright"){
                bottom = bottom + 1
              }
}
df = tibble(MeanL = mean(data$LoserLegs),
            VariaceL = var(data$LoserLegs),
            Top = top,
            Bottom = bottom)
```

## Altham's Multiplicative Binomial Distribution
For each player in the 2024 season, use the number of legs won in each match to compute maximium likelihood estimates for $p$ and $\theta$. \
**Note**: The "for each" here can be done more efficiently with a function or a loop!

* Create a vector containing the number of legs won by the player. Note sometimes you will need to pull from winner's legs and sometimes from the loser's legs.
* Find the MLE for the player using this data. Ensure to report $p$ and $\theta$ for each player in a nicely formatted table.\
**Note:** In Lecture on Monday, we used transformations to restrict how `optim()` searches the parameter space. For example, here, you might consider
`p <- 1/(1+exp(-par[1]))` and `theta <- exp(par[2])`.
* Make comments about the MLEs in the context of the standings.

**Solution:**
Typicially, the higher the player's ranking the higher their p values of probability to win are. There is one outlier being that Nathan Aspinall has a high p value without being a top player. The reverse is true about the theta values, they decrease with higher player rankings.
```{r}
total_legs = function(player, data){
        winner_legs = data |> 
        filter(Winner == player) |> 
        pull(WinnerLegs)
        loser_legs = data |> 
        filter(Loser == player) |> 
        pull(LoserLegs)
        c(winner_legs, loser_legs)
}

players = c("Luke Littler","Luke Humphries","Michael van Gerwen",
             "Michael Smith","Nathan Aspinall","Rob Cross",
             "Gerwyn Price","Peter Wright")

negative_log <- function(par, x){
        p = 1/(1+exp(-par[1]))   
        theta = exp(par[2])      
        n = 6  
        loglik = sum(log(choose(n, x)) + x*log(p) + (n-x)*log(1-p) + theta * x*(n-x))
        -loglike_altbinom(x, n, p, theta)
}

result = tibble(Player=players, p=NA, theta=NA)

for (i in seq_along(players)) {
      x = total_legs(players[i], data)
      fit = optim(c(0,0), negative_log, x=x)
      result$p[i] = 1/(1+exp(-fit$par[1]))
      result$theta[i] = exp(fit$par[2])
}

result
```

## Altham's Multiplicative Beta Binomial Distribution

For each player in the 2024 season, use the number of legs won in each match to compute maxmium likelihood estimates for $p$ and $\theta$.\
**Note**: The "for each" here can be done more efficiently with a function or a loop!

* Create a vector containing the number of legs won by the player. Note sometimes you will need to pull from winner's legs and sometimes from the loser's legs.
* Find the MLEs for the player using this data. Ensure to report $\alpha$, $\beta$, and $\theta$ for each player in a nicely formatted table.\
**Note 1**: Due to the computational complexity of `integrate()`, you should compute the logged probability for `0:6` once and add them according to the data in each player's vector.
**Note 2**: In Lecture on Monday, we used transformations to restrict how `optim()` searches the parameter space.
* Plot the underlying Beta($\alpha$, $\beta$) distribution for each player. \
**Note**: Add the mean and variance of the Beta to the table of MLEs. Recall
\begin{align*}
 E(X)   &= \frac{\alpha}{\alpha+\beta}\\
 sd(X) &= \sqrt{\frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}}
\end{align*}
* Make comments about the MLEs in the context of the standings.\
**Note**: The players alternate who throws first in a leg has the mathematical "head start" to reach a finish before their opponent can take another turn. Players alternate who throws first with each leg.


**Solution:**
```{r}
#| echo: false
loglike_altbb = function(x, size, alpha, beta, theta) {
        probs = daltbbinom(x, size, alpha, beta, theta)
        probs[probs < 1e-10] = 1e-10
        sum(log(probs))
}

neg_loglike_altbb = function(par, x, size = 6) {
        alpha = exp(par[1])
        beta  = exp(par[2])
        theta = exp(par[3])
        -loglike_altbb(x, size, alpha, beta, theta)
}

results_altbb = tibble(
        Player = players,
        alpha = NA,
        beta = NA,
        theta = NA,
        MeanBeta = NA,
        SDBeta = NA
)

for (i in seq_along(players)) {
        x = total_legs(players[i], data)
        fit = optim(c(log(1), log(1), log(1)), neg_loglike_altbb, x = x, method="Nelder-Mead")
  
        alpha_hat = exp(fit$par[1])
        beta_hat  = exp(fit$par[2])
        theta_hat = exp(fit$par[3])
  
        results_altbb$alpha[i] = alpha_hat
        results_altbb$beta[i]  = beta_hat
        results_altbb$theta[i] = theta_hat
        results_altbb$MeanBeta[i] = alpha_hat / (alpha_hat + beta_hat)
        results_altbb$SDBeta[i]   = sqrt(alpha_hat * beta_hat / ((alpha_hat + beta_hat)^2 * (alpha_hat + beta_hat + 1)))
}

results_altbb
```

## Executive Summary
Summarize the work you've done here to provide a brief (200-300 word) data-driven summary of the 2024 Premier League season using your work in Lab 10. Your summary should be professional, clear, and all claims should be corroborated with statistical support.

**Solution:**
Although the premier league only has 8 contestants, there is a large gap between the top players and bottom players. Even with this gap, losing players still won an average of approximately 3.56 legs, meaning that most matches were competitive rather than a blow-out. The number of legs won isn't as skewed as the win count between top and bottom which supports that the games were "close" at the least.

Altham’s Multiplicative Binomial model showed maximum likelihood estimates that higher-ranked players tend to have larger values of $p$, indicating a higher probability of winning singular legs. Those players generally exhibit lower values of $theta$, which indicates their more consistent performance. In contrast, lower ranked players have smaller $p$ values and often higher theta which reflectes lower success rates and greater variability, though the difference isn't substantial.

The Multiplicative Beta Binomial model adds the variability in leg winning ability compared to the Altham's model. Players with higher estimated Beta means align with the top of the rankings, which confirmes stronger average performance. Larger Beta standard deviations and higher $theta$ values among lower ranked players indicate more inconsistency.

Overall, the modeling supports the standings pf the leader board stronger players consistently win more legs and exhibit more stable performance, while weaker players win fewer legs and display greater variability, showing a clear skill gap across the league.
x