---
title: Lab Eleven -- Point Estimation and Resampling # Title of the document

# Bibliography Details
bibliography: Lab11References.bib  # Name of bib file in the same folder 
bibliographystyle: apa            # Bibliography style

execute:               # Default settings for code blocks
  warning: false       # Do not print warnings
  message: false       # Do not print messages
  size: "scriptsize"   # Set code size to script size

# Below will do html by default. To render the entire project
# Use the following code:
# quarto::quarto_render("Lab10WriteUp.qmd")
format:
  pdf:        # Specify .pdf formatting options 
    documentclass: article     # Document Class
    number-sections: true      # Number Sections
    reference-section-title: "References" # Title for reference section
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables 
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in
  html:       # Specify HTML formatting options 
    number-sections: true      # Number Sections
    code-fold: true            # Button to show code
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables and figures
format-links: [pdf]
---
```{r}
#| echo: false

default_chunk_hook  <- knitr::knit_hooks$get("chunk")

latex_font_size <- c("Huge", "huge", "LARGE", "Large", 
                     "large", "normalsize", "small", 
                     "footnotesize", "scriptsize", "tiny")

knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  if(options$size %in% latex_font_size) {
    paste0("\n \\", options$size, "\n\n", 
      x, 
      "\n\n \\normalsize"
    )
  } else {
    x
  }
})
```

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
library(patchwork)
```

**Tips**

* Complete the tasks below. Make sure to start your solutions in on a new line that starts with "**Solution**:".
* Make sure to use the Quarto Cheatsheet. This will make completing and writing up the lab *much* easier.
* Make sure to use "enter" to make your code readable, so it doesn't run off the page. I switched the Quarto document to automatically render to .pdf, so you can see what you're submitting by default. You can switch back for the highlighting by moving the html block above the pdf block in the YAML header.
* Don't forget that you can copy-and-paste code when you're doing something similar as we did in class!
* Make sure to plot all computed probabilities. This is good `ggplot` practice AND will help make sure you don't make a mistake when computing the probability.


# Introduction
When performing point estimation, there isn't just one *right* answer. For example, the Method of Moments (MOM) and Maximum Likelihood Estimates (MLEs) aren't always the same.

One question that comes up, is why we calculate the sample variance as
$$S_{n-1}^2 = \frac{1}{n-1} \sum_{i=1}^n (X_i -\bar{X})^2,$$
when it is supposed to be the ``average squared distance from the mean, i.e., 
$$S_n^2 = \frac{1}{n} \sum_{i=1}^n (X_i -\bar{X})^2.$$


# Comparing Estimators

## Initial Simulation
In this lab, we will explore why we use $S_{n-1}^2$ and not $S_n^2$. To do this, we will generate data from four known distributions:

* Poisson$(\lambda=2)$
* Binomial$\left(n=50, p=\frac{5-\sqrt{21}}{10}\right)$
* Normal$(\mu=0, \sigma=\sqrt{2})$
* Exponential$\left(\lambda = \frac{1}{\sqrt{2}}\right).$

**Remark**: Note that I have selected the parameters so that the true variance for all four models is 2.

For each distribution, write a loop that completes the following 1000 times. Put `set.seed(7272)` at the top of your code chunk.

* Generate a sample of size $n$ from the distribution.
* Calculate and save $S_{n-1}^2$ and $S_n^2.$

Start with $n=20$. Is there evidence that $S_{n-1}^2$ is a better estimator than $S_n^2$?

Create a plot or combination of plots (using patchwork) to show the results across all four distributions. Make sure to create a corresponding table that numerically summarizes the results of the simulation. You should include
\begin{align*}
\textrm{Bias} &= \textrm{Average of Estimates} - \textrm{True Value}\\
\textrm{Precision} &= \frac{1}{\textrm{Variance of Estimates}}\\
\textrm{Mean Square Error} &= \textrm{Bias}^2 + \textrm{Variance of Estimates}.
\end{align*}

**Solution:**
```{r}
#| echo: false
set.seed(7272)
n = 20
sim = 1000
sample_v1 = numeric(sim)
sample_v = numeric(sim)
true = 2

# Poisson lambda = 2
lambda = 2
for (i in 1:sim){
    distP = rpois(n, lambda)
    sample_v1[i] = var(distP)
    sample_v[i] = mean((distP - mean(distP))^2)
}
bias1 = round(mean(sample_v1) - true, 4)
variance1  = round(var(sample_v1), 4)
precision1 = round(1 / variance1, 4)
mse1 = round(bias1^2 + variance1, 4)

bias = round(mean(sample_v) - true, 4)
variance  = round(var(sample_v), 4)
precision = round(1 / variance, 4)
mse = round(bias^2 + variance, 4)

dfP = tibble(
Distribution = "Poisson",
    BiasN_1 = bias1,
    VarianceN_1 = variance1,
    PrecisionN_1 = precision1,
    MSEN_1 = mse1,
    Bias = bias,
    Variance = variance,
    Precision = precision,
    MSE = mse
)
valuesP = sample_v
valuesP1 = sample_v1


# Binomial n = 50, p = (5 - root21)/10
B_n = 50
p = (5 - 21^(1/2))/10
for (i in 1:sim){
    distB = rbinom(n, B_n, p)
    sample_v1[i] = var(distB)
    sample_v[i] = mean((distB - mean(distB))^2)
}
bias1 = round(mean(sample_v1) - true, 4)
variance1  = round(var(sample_v1), 4)
precision1 = round(1 / variance1, 4)
mse1 = round(bias1^2 + variance1, 4)

bias = round(mean(sample_v) - true, 4)
variance  = round(var(sample_v), 4)
precision = round(1 / variance, 4)
mse = round(bias^2 + variance, 4)

dfB = tibble(
Distribution = "Binomial",
    BiasN_1 = bias1,
    VarianceN_1 = variance1,
    PrecisionN_1 = precision1,
    MSEN_1 = mse1,
    Bias = bias,
    Variance = variance,
    Precision = precision,
    MSE = mse
)
valuesB = sample_v
valuesB1 = sample_v1


# Normal mu = 0, sigma = root2
mu = 0
sigma = 2^(1/2)
for (i in 1:sim){
    distN = rnorm(n, mu, sigma)
    sample_v1[i] = var(distN)
    sample_v[i] = mean((distN - mean(distN))^2)
}
bias1 = round(mean(sample_v1) - true, 4)
variance1  = round(var(sample_v1), 4)
precision1 = round(1 / variance1, 4)
mse1 = round(bias1^2 + variance1, 4)

bias = round(mean(sample_v) - true, 4)
variance  = round(var(sample_v), 4)
precision = round(1 / variance, 4)
mse = round(bias^2 + variance, 4)

dfN = tibble(
Distribution = "Normal",
    BiasN_1 = bias1,
    VarianceN_1 = variance1,
    PrecisionN_1 = precision1,
    MSEN_1 = mse1,
    Bias = bias,
    Variance = variance,
    Precision = precision,
    MSE = mse
)
valuesN = sample_v
valuesN1 = sample_v1


# Exponential lambda = 1/root2
lambda = 1/(2^(1/2))
for (i in 1:sim){
    distE = rexp(n, lambda)
    sample_v1[i] = var(distE)
    sample_v[i] = mean((distE - mean(distE))^2)
}

bias1 = round(mean(sample_v1) - true, 4)
variance1  = round(var(sample_v1), 4)
precision1 = round(1 / variance1, 4)
mse1 = round(bias1^2 + variance1, 4)

bias = round(mean(sample_v) - true, 4)
variance  = round(var(sample_v), 4)
precision = round(1 / variance, 4)
mse = round(bias^2 + variance, 4)

dfE = tibble(
Distribution = "Exponential",
    BiasN_1 = bias1,
    VarianceN_1 = variance1,
    PrecisionN_1 = precision1,
    MSEN_1 = mse1,
    Bias = bias,
    Variance = variance,
    Precision = precision,
    MSE = mse
)
valuesE = sample_v
valuesE1 = sample_v1
```

```{r}
#| echo: false
# Plotting
table = bind_rows(dfP, dfB, dfN, dfE) |>
          pivot_longer(
                cols = -Distribution,
                names_to = c("Metric", "EstimatorFlag"),
                names_sep = "N",  # splits Bias1 vs Bias etc.
                values_to = "Value") |>
          mutate(
                Estimator = ifelse(!is.na(EstimatorFlag), "{n-1}²", "n²")) |>
          select(-EstimatorFlag)
table |>
        pivot_wider(names_from = Metric, values_from = Value) |>
        knitr::kable(caption = "Simulation Summary of Variance Estimators")

plot = tibble(
          Estimate = c(valuesP1, valuesP, valuesB1, valuesB, valuesN1, valuesN, valuesE1, valuesE),
          Estimator = rep(c("{n-1}²", "n²"), times = 4, each = 1000),
          Distribution = rep(c("Poisson", "Binomial", "Normal", "Exponential"), each = 1000 * 2)
)
ggplot(plot, aes(x = Estimate, fill = Estimator)) +
        geom_density(alpha = 0.5) +
        geom_vline(xintercept = 2, linetype = "dashed", color = "black") +
        facet_wrap(~Distribution, scales = "free") +
        labs(
            title = "Comparison of Variance Estimators",
            subtitle = "True Variance = 2",
            x = "Estimated Variance",
            y = "Density",
            fill = "Estimator") +
        theme_minimal()
```

## Simulation over various sample sizes
Consider the effect of the sample size have on these results. Provide a graphical and numerical summary of what happens as $n$ increases -- say $n=20$, $n=50$, $n=100$, and $n=500$. Does the result in your initial simulation hold across samples? Or, do things change as the sample size changes? Ensure to set your randomization seed.

**Solution:**
```{r}
#| echo: false
set.seed(7272)

sample_sizes = c(20, 50, 100, 500)
sim = 1000
true = 2

results = list()

for (n in sample_sizes) {
  
  compute = function(x) {
    v_n1 = var(x)
    v_n = mean((x - mean(x))^2)
    return(c(v_n1, v_n))
  }
  
  run = function(generator) {
    out = replicate(sim, {
      x = generator(n)
      compute(x)
    })
    t(out)
  }
  
  pois = run(function(n) rpois(n, 2))
  bino = run(function(n) rbinom(n, 50, (5 - sqrt(21))/10))
  norm = run(function(n) rnorm(n, 0, sqrt(2)))
  expo = run(function(n) rexp(n, 1/sqrt(2)))
  
  dists = list(Poisson = pois, Binomial = bino, Normal = norm, Exponential = expo)
  
  for (name in names(dists)) {
      vals = dists[[name]]
    
      v_n1 = vals[,1]
      v_n = vals[,2]
    
      df = tibble(
              n = n,
              Distribution = name,
              Estimator = c("{n-1}²", "n²"),
              Bias = c(mean(v_n1) - true, mean(v_n) - true),
              Variance = c(var(v_n1), var(v_n))) |>
            mutate(Precision = 1 / Variance,
                  MSE = Bias^2 + Variance)
    
      results[[length(results) + 1]] = df
  }
}

table = bind_rows(results)
wide = table |>
  pivot_wider(
    names_from = Estimator,
    values_from = c(Bias, Variance, Precision, MSE)
  )

```
```{r}
#| echo: false
ggplot(table, aes(x = n, y = MSE, color = Estimator)) +
      geom_line() +
        geom_point() +
      facet_wrap(~Distribution) +
      labs(title = "MSE vs Sample Size") +
      theme_minimal()

wide |>
  knitr::kable(
    caption = "Simulation Results Across Sample Sizes",
    digits = 4,
    col.names = c(
      "n", "Distribution",
      "Bias (n-1)", "Bias (n)",
      "Var (n-1)", "Var (n)",
      "Precision (n-1)", "Precision (n)",
      "MSE (n-1)", "MSE (n)"
    )
  )
```

## Simulation
You have written code to draw a random sample of various sizes from each distribution, and to calculate and store the calculated estimates. Rework your code to produce a density plot of the $s_{n-1}^2$ and $s_{n}^2$. Do you have a guess about their distribution? Ensure to set your randomization seed. 

**Solutution:**
As the sampling size increases, the distributions appear to show signs of approaching the Normal distribution, and while they are small sample sizes they have a right skew. This could resemble a log-normal distribution since at low sample sizes it shows large right skews. Since the variances of $(n-1)^2$ and $n^2$ are so close to each other, it is incredibly hard to distinguish them on the grpahs.
```{r}
#| echo: false
set.seed(7272)

plot2 = list()

compute = function(x) {
  v_n1 = var(x)
  v_n = mean((x - mean(x))^2)
  return(c(v_n1, v_n))
}

run = function(generator, n) {
  out = replicate(sim, {
    x = generator(n)
    compute(x)
  })
  t(out)
}

for (n in sample_sizes){
  
      pois = run(function(n) rpois(n, 2), n)
      bino = run(function(n) rbinom(n, 50, (5 - sqrt(21))/10), n)
      norm = run(function(n) rnorm(n, 0, sqrt(2)), n)
      expo = run(function(n) rexp(n, 1/sqrt(2)), n)
  
      dists = list(Poisson = pois, Binomial = bino, Normal = norm, Exponential = expo)
  
      for (name in names(dists)) {
        vals = dists[[name]]
    
        df = tibble(Estimate = c(vals[,1], vals[,2]),
                  Estimator = rep(c("{n-1}²", "n²"), each = sim),
                  Distribution = name,
                  n = n)
    
        plot2[[length(plot2) + 1]] = df
  }
}

plotDF = bind_rows(plot2)
plotDF = plotDF |>
        mutate(Distribution = ifelse(Distribution == "Exponential", "Expo.", Distribution))
ggplot(plotDF, aes(x = Estimate, fill = Estimator)) +
        geom_density(alpha = 0.5) +
        facet_grid(Distribution ~ n, scales = "free") +
        geom_vline(xintercept = 2, linetype = "dashed") +
        labs(title = "Density of Variance Estimators",
            x = "Estimated Variance",
            y = "Density",
            fill = "Estimator") +
        theme_minimal()
```

## Bootstrap
Now, take *one* sample from each distribution. Instead of drawing a random sample at each iteration, sample *with replacement* from the one sample. Produce a density plot of the $s_{n-1}^2$ and $s_{n}^2$. Do we get roughly the same information as we do in the full simulation? Try this over several seeds -- what do you notice? Ensure to set your randomization seed.

**Solution:***
For the most part it is close enough to see similarities between the bootstrapping and the full simulation, showing similar curves. It still shows a general right skew, however across different seeds the distributions can get extremely incorrect and far off from where the full simulation shows.
```{r}
#| echo: false
set.seed(7272)

n = 20
B = 1000

bootstrap_run = function(x) {
      compute = function(xb) {
            v_n1 = var(xb)
            v_n = mean((xb - mean(xb))^2)
            c(v_n1, v_n)
      }
  
      out = replicate(B, {
            xb = sample(x, size = length(x), replace = TRUE)
            compute(xb)})
  
      t(out)
}

p1 = rpois(n, 2)
b1 = rbinom(n, 50, (5 - sqrt(21))/10)
n1 = rnorm(n, 0, sqrt(2))
e1 = rexp(n, 1/sqrt(2))

pois = bootstrap_run(p1)
bino = bootstrap_run(b1)
norm = bootstrap_run(n1)
expo = bootstrap_run(e1)

plotDF = bind_rows(
  tibble(Estimate = c(pois[,1], pois[,2]),
         Estimator = rep(c("(n-1)", "n"), each = B),
         Distribution = "Poisson"),
  
  tibble(Estimate = c(bino[,1], bino[,2]),
         Estimator = rep(c("(n-1)", "n"), each = B),
         Distribution = "Binomial"),
  
  tibble(Estimate = c(norm[,1], norm[,2]),
         Estimator = rep(c("(n-1)", "n"), each = B),
         Distribution = "Normal"),
  
  tibble(Estimate = c(expo[,1], expo[,2]),
         Estimator = rep(c("(n-1)", "n"), each = B),
         Distribution = "Exponential")
)

ggplot(plotDF, aes(x = Estimate, fill = Estimator)) +
      geom_density(alpha = 0.5) +
      facet_wrap(~Distribution, scales = "free") +
      geom_vline(xintercept = 2, linetype = "dashed") +
      labs(
          title = "Bootstrap Sampling Distributions",
          x = "Estimated Variance",
          y = "Density",
          fill = "Estimator") +
      theme_minimal()
```


# Executive Summary
Summarize the work you've done here to provide a brief (200-300 word) summary of the simulation and the results using your work in Lab 11. Your summary should be professional, clear, and all claims should be corroborated with statistical support.

**Solution:**
This simulation lab showed why the sample variance is calculated with a divisor of $(n−1)$ instead of $n$ to remove bias. We compared both variance estimators across four the distributions of Poisson, Binomial, Normal, and Exponential, using a set true variance of 2. At each distribution we used repeated sampling scaling in size to measure bias, variance, precision, and mean squared error.

Across all distributions, the $(n−1)$ version of the sample variance performed better because it stayed closer to the true variance on average. However, the version that divides by $n$ consistently underestimated the true variance which showed a clear negative bias. This pattern was the same for every distribution, suggesting the result is general and not dependent on a specific model.

As sample size increased both estimators became more stable. Their variability decreased, and their sampling distributions became more concentrated around the true value of 2. For small samples, the distributions were right skewed, but they became more symmetric as sample size grew, approaching a normal distribution. At large sample sizes, the two estimators became nearly indistinguishable, with just slight discrepancies around their peaks. The bootstrap results that resampled from a single dataset, showed similar overall patterns but were more sensitive to the particular sample used. Different random seeds led to slightly different shapes.

Overall, the $(n−1)$ version of the sample variance is preferred because it removes bias and provides more accurate estimates across all distributions and sample sizes studied.