---
title: Lab One -- Installing R, Positron, and tinytex # Title of the document

# Bibliography Details
bibliography: Lab1References.bib  # Name of bib file in the same folder 
bibliographystyle: apa                   # Bibliography style

execute:
  warning: false
  message: false

# Below will do html by default. To render the entire project
# Use the following code:
# quarto::quarto_render("[FILENAME].qmd")

format:
  html: 
    number-sections: TRUE
    code-fold: TRUE
    code-line-numbers: TRUE
    cap-location: top 
  pdf:
    documentclass: article
    number-sections: TRUE
    reference-section-title: "References" 
    code-line-numbers: TRUE
    cap-location: top 
    include-in-header:
      - text: |
          \usepackage{amsmath}
          \usepackage[dvipsnames]{xcolor}
          \usepackage[shortlabels]{enumitem}
          \usepackage{parskip}
          \usepackage{hyperref}
          \usepackage{geometry}
          \usepackage{float}
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in
format-links: [pdf]
---

```{r}
#| echo: false

default_chunk_hook  <- knitr::knit_hooks$get("chunk")

latex_font_size <- c("Huge", "huge", "LARGE", "Large", 
                     "large", "normalsize", "small", 
                     "footnotesize", "scriptsize", "tiny")

knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  if(options$size %in% latex_font_size) {
    paste0("\n \\", options$size, "\n\n", 
      x, 
      "\n\n \\normalsize"
    )
  } else {
    x
  }
})

```

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
library(palmerpenguins)
```

# Install Software

-   [Install R](https://cran.r-project.org/).
-   [Install Positron](https://positron.posit.co/download.html).
-   [Install tinytex](https://yihui.org/tinytex/).

# Set Up GitHub

-   [Sign Up for GitHub](https://github.com/).
-   Sign Up for [Our GitHub Classroom](https://classroom.github.com/classrooms/156721198-colgateuniversitycipolli-sp2026-math240).
-   Pull the GitHub Repo for Lab One [here](https://classroom.github.com/a/8jPlfX90).

# Compile the Lab Template

-   Compile the Lab One template as an html file (default).
-   Use the commented code to compile the Lab One template as a pdf file.

# Learn to Use Quarto and R

-   Recreate this section in your **`Lab1WriteUp.qmd`** file.

## Bold, Italics, Underline

**This is bold text**

*This is italicized text*

<u>This is underlined text</u>

## Lists and Enumerated Lists

-   This is item one
    -   This is item a
    -   This is item b
-   This is item two
-   This is item three
    -   This is item c
    -   This is item d
    -   This is item e
-   This is item four

1.  This is item one
    a.  This is item a
    b.  This is item b
    c.  This is item c
2.  This is item two
3.  This is item three
4.  This is item four

## Math

The quadratic formula
\begin{align*}
x = \frac{-b \pm \sqrt{b^2-4ac}}{2a}.
\end{align*}
It is helpful for solving equations like
\begin{align*}
x^2+2x-13=0.
\end{align*}
Using $a=1,b=2,c=-13,$ we find solutions using the quadratic formula.
\begin{align*}
x &= \frac{-b \pm \sqrt{b^2-4ac}}{2a} \tag{Definition} \\
&= \frac{-(2) \pm \sqrt{(2)^2-4(1)-13}}{2(1)} \tag{Definition} \\
&= \frac{-2 \pm \sqrt{56}}{2} \tag{Symplifying} \\
&= \frac{-2}{2} \pm \frac{\sqrt{56}}{2} \tag{Rewriting} \\
&= \frac{-2}{2} \pm \sqrt{\frac{56}{4} \tag{Rewriting} }\\
&= -1 \pm \sqrt{14} \tag{Symplifying} \\
&= -4.741657 \text{ or } 2.741657
\end{align*}

## Figures

We can *show* we've found the correct answer by plotting the line and the points (see @fig-one-quadratic).
```{r}
#| label: fig-one-quadratic
#| fig-cap: "Graph of the quadratic."
#| fig-width: 5
#| fig-height: 3
#| scale: "80%"
#| fig-align: "center"
#| size: scriptsize 
dat <- tibble(x=seq(-6, 6, length.out=1000))|>
              mutate(y = x^2 + 2*x -13)

ggplot(data=dat)+
      geom_line(aes(x=x, y=y))+
      geom_hline(yintercept=0, alpha=0.50)+
      geom_vline(xintercept=0, alpha=0.50)+
      theme_bw()+
      labs(x="x", y="y")+
      annotate("point", x=-1-sqrt(14), y=0, color="red")+
      annotate("point", x=-1+sqrt(14), y=0, color="red")
```

## Tables

Note, we can pick points to show intermediate value theorem, noting that our function is continuous. For example, in @tbl-points, we can see there must be a root between $x = -5$ and $x = 0$ and another between $x=0$ and $x=5$.
```{r}
#| label: tbl-points
#| tbl-cap: "Table of values of the quadratic."
#| tbl-align: "center"
#| size: scriptsize
tibble(x = c(-5, -1-sqrt(14), 0, -1+sqrt(14), 5)) |>
      mutate(y = x^2 + 2*x -13)|>
      knitr::kable(digits=4)
```

## Citations

Above, I used the `tidyverse` package [@tidyverse] to create @fig-one-quadratic and @tbl-points. Additionally, I used the `knitr` package [@knitr] to nicely format the table.

Note that you can ask for citations for **`R`** packages using the **`citation()`** function.

## Delivery

Ensure to render the full project using the following code, so the PDF link will work and a .pdf file will be created.


## References