---
title: Tidyverse and GGplot Guide # Title of the document

# Bibliography Details
bibliography: Tidyverse-GGplot-Guide.bib  # Name of bib file in the same folder 
bibliographystyle: apa            # Bibliography style

execute:               # Default settings for code blocks
  warning: false       # Do not print warnings
  message: false       # Do not print messages
  size: "scriptsize"   # Set code size to script size

# Below will do html by default. To render the entire project
# Use the following code:
# quarto::quarto_render("Lab10WriteUp.qmd")
format:
  pdf:        # Specify .pdf formatting options 
    documentclass: article     # Document Class
    number-sections: true      # Number Sections
    reference-section-title: "References" # Title for reference section
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables 
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in
  html:       # Specify HTML formatting options 
    number-sections: true      # Number Sections
    code-fold: true            # Button to show code
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables and figures
format-links: [pdf]
---

```{r}
#| echo: false

default_chunk_hook  <- knitr::knit_hooks$get("chunk")

latex_font_size <- c("Huge", "huge", "LARGE", "Large", 
                     "large", "normalsize", "small", 
                     "footnotesize", "scriptsize", "tiny")

knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  if(options$size %in% latex_font_size) {
    paste0("\n \\", options$size, "\n\n", 
      x, 
      "\n\n \\normalsize"
    )
  } else {
    x
  }
})
```

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
library(patchwork)
library(pwr)
library(readxl)
library(dplyr)
library(kableExtra)
```



# Study Introduction
The dataset used in this guide comes from the published study Skeletal Correlates for Body Mass Estimation in Modern and Fossil Flying Birds by Daniel J. Field, Colton Lynner, Cheryl Brown, and Simon A. F. Darroch, published in PLoS One in 2013. Their purpose was to examine how skeletal size relates to body mass in certain birds. They measured various bone lengths in hundreds of birds to determine which bones best predict the species and size of the bird they came from. This type of data is useful because body mass is an important distinguishing trait, and the relationship between mass and bone size can also help scientists estimate the mass of extinct birds whose only remains are skeletons. This guide will explore obtaining data from a published article, cleaning data using the tidyverse package, and plotting the data to identify similarities and differences among different bird species.

## Obtaining Data
Before we can work with the data to determine important differences between the bird species, we must first collect the data into a readable format.

**Find and Create the CSV:**
- Find the article at: https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0082000
- Scroll down to "*Supporting Information*" and click on Table S1. This will download a CSV of the data
- Rename and save the CSV, then move it into the same folder as your project


## Starting Tidyverse and Data Cleaning
Now that we have access to the data in a readable form, we must put it into a dataframe so that we can easily clean and re-organize the data to our liking.

**Start Tidyverse:**

- Install and load the `tidyverse` package into `R` so that we can access its funcitons

```{r}
# install.packages("tidyverse")   ## Installs the package
library(tidyverse)  ## Loads the package
```

- Read the CSV file into a tibble dataframe to manipulate later

```{r}
dat.Birds = read_csv("Table_S1.csv")   ## read_csv() makes a tibble
```

**Cleaning the Data:**

- We will rename the columns to not have `" "`s so that they are easier to call later on

- Luckily the researchers didn't record bad data with "`NA`"s in it, but generally we should remove them

- We will remove all other columns that won't be needed for our purposes

- We will use the `|>` operator to pass the previous item in as the first argument of the next call to save time

```{r}
dat.Birds.cleaned = dat.Birds |>       ## Reset dat.Birds to clean version and pass it through
            rename_with(~ gsub(" ", ".", .x)) |>    ## Replacing " "  with "."
            rename(mass_g = 'mass.(g)') |>      ## Making a cleaner name
            rename(femur_circ = 'femur_circumference.(mm)') |>
            rename(order = Order) |>
            select(order, species, sex, femur_circ, mass_g) |>  ## Keeping relevant columns only
            drop_na()             ## Removing all rows with NA's
```


## Creating Plots with ggplot2
Now we will use `ggplot2` to create visualizations of the data to understand their shape and structure. We will need to pull certain variables and pick the correct graph depending on the type of data we use.

**Data Types and Graphs:**

1. Categorical data : Bar plot (geom_bar())

  - total birds observed in each order

2. Quantitative (single variable) : Histogram (geom_histogram())

  - distribution of bird's body mass

3. Quantitative vs Categorical : Boxplot (geom_boxplot())

  - body mass of female vs. male birds

4. Quantitative vs Quantitative : Scatterplot (geom_point())

  - femur size vs. body mass

**Creating Plots:**

